\chapter{Conclusions and Outlook}

The first conclusion, although it is actually almost a premise to this dissertation, is that sputtering is indeed an important effect that needs considering in high-fluence ion irradiation. This is especially true in the irradiation of nanostructures, where sputter yields can be much larger than in bulk. The diameter dependent enhancement of the sputter yield was shown for $Si$-nanowires in this thesis. These results show that a good qualitative estimation, or intuition, of how any given nanostructure will be sputtered can be obtained by using the Sigmund model for sputtering. The relative size of the overlap of the ions' nuclear energy loss and the surface of the target, even if it is nanostructured, is a reasonable estimation for the relative sputter yield. Thus, a feeling for which part of a complicated nanostructure will be most affected by sputtering during the irradiation with ions of a certain species and energy can be gained. For a specific nanoscale geometry the MC BCA simulation program \emph{iradina} \cite{borschel_ion_2011} can be used to make a more detailed analysis. The diameter dependence of the sputter yield simulated with \emph{iradina} is qualitatively reproduced by the experiments reported in this thesis.

The quantitative values of sputtering are, in general, not accessible through the naive use of MC BCA simulations. However, very good agreement can be found for certain material and ion combinations \cite{biersack_computer_1987,hofsass_simulation_2014}, so the situation is not at all hopeless. The main difficulty is to find correct low energy interaction potential for the colliding atoms and ions for the given situation. Because the secondary ion mass spectrometry (SIMS) technique is highly reliant on sputter yields and MD simulations also require the correct interaction potential at low particle energies, there is some interest in solving this problem \cite{nordlund_interatomic_2008}. Because sputtering is dominated by low energy collisions, it is very sensitive to the interaction potential precisely at the energy range where it is not easily accessible to other experiments, while the available theoretical models differ. An example of interesting sputtering behavior is found in $Ni$ and $Cu$, which are very similar in mass and density, yet show a factor of two difference in their sputter yield, because of differences in surface binding, electronic energy loss and nuclear interaction \cite{biersack_computer_1987}. 

Experiments on sputtering of defined nanostructures, such as the ones performed on nanowires within this thesis, may be a useful approach to test theoretical predictions based on different interaction potentials. This approach is not limited to semiconductor nanowires, only by the availability of the nanowires of the desired material. This is not a significant constraint, because in addition to the bottom-up growth methods, template driven nanowire synthesis methods are very versatile \cite{martin_membrane-based_1996}. The systematical deviation, such as the one caused in the experiments reported here by oxidation of the $Si$-nanowires can certainly be controlled by an optimized experimental protocol. Overall, a quantitative evaluation of the sputter yield should be possible. Such experiments should be combined and compared with ion impact-angle dependent measurements of the sputtering \cite{hofsass_simulation_2014} and the angle resolved emission of the sputtered atoms \cite{wirtz_storing_2008,verdeil_angular_2008}. The combination of such experiments can not produce the correct interaction potential, however, it can be a powerful test for results from simulations with different interaction potentials to determine which describes the inter-atomic interaction best.

The main goal of ion irradiation is typically not sputtering, but the incorporation of dopants in the target. For nanostructured targets, care has to be taken to avoid an inhomogeneous irradiation and doping profile, which can be caused by shadowing of the ion beam in a nanostructure or amongst separate nanostructures. This was illustrated with the nano-XRF investigation of $Mn$-doped $ZnO$ nanowires. The nano-XRF investigation shows that the BC MCA simulation is adequate for the prediction of the doping concentration for low ion fluences in nanowires. The limit of this applicability is given by the point where around $20\%$ of the material affected by the ion beam is sputtered, which in nanostructures is typically $20\%$ of the whole nanostructure's volume. Similar approximations can be made in bulk \cite{moller_tridyn_1984,andersen_computer_1986,moller_tridyn-binary_1988,sigmund_alloy_1993,zaporozchenko_preferential_1995}, even if the given references don't explicitly state a limiting ion fluence or sputtered depth. For ion irradiation with higher ion fluences, dynamic simulations are needed to predict the correct dopant concentration and profile. The limiting ion fluence can be much lower for nanostructures than in bulk, because there is less material to be sputtered and sputtering is enhanced. For high fluence ion irradiation, where more than $20\%$ of the material is expected to be sputtered, dynamic simulations are recommended. Software, which can dynamically change the structure and composition of the ion irradiated, nanostructured target, has been revealed recently in reference \cite{moller_tri3dyn_2014}. A comparison of the experimental results presented in this thesis with the results from such a simulation is a logical next step.

Down these lines, the application of the nano-XRF quantification technique to ion irradiated nanostructures can produce further interesting results. Because nano-XRF is highly sensitive to elemental concentrations, it can widen the scope of the proposed studies into sputtering by investigating compositional changes in ion irradiated nanostructures of compound materials. In compound materials preferential sputtering of one of the materials' components may become relevant even before a high dopant concentration has been reached. The interplay of nano-structuring, compositional changes and preferential sputtering could thus be investigated for a vast array of materials, by no means limited to semiconductors. A comparison to simulation results would further increase the understanding of the parameters influencing the preferential sputtering, which has practical meaning in secondary ion mass spectroscopy (SIMS), but also in the development of materials for fusion reactor components \cite{kelly_attempt_1978,roth_sputtering_1990,kenmotsu_effect_2011}. Using nanowires for such an experiment has the advantage that samples with multiple diameters can be fabricated, irradiated and investigated in parallel, and thus a larger parameter space becomes accessible to simultaneous investigation.

The sputtering of nanostructures is enhanced relative to bulk, not only because of the high surface to volume ratio, but also by thermal effects. These can be very pronounced if the energy deposited by the impinging ion is confined to a small volume \cite{greaves_enhanced_2013,ilinov_sputtering_2014,nietiadi_sputtering_2014,anders_sputtering_2015,urbassek_sputter_2015}. This can lead to explosive ejection of large clusters of $1000s$ of atoms. Such extreme thermal sputtering effects were not observed in the experiments presented in this thesis; firstly, because the nanowires are relatively large compared to the ion energy deposited in them, leading, on average, to only little energy deposited per atom; secondly, the ion mass and target atom mass are both relatively low, leading to a low stopping power and a low density of the energy deposition. The simulation of sputtering with the BCA and the Sigmund theory would break down in experiments where this is not the case. Nevertheless, the maximum sputter yield is observed at lower nanowire diameters in the experiment than in the simulation. This could be caused by the correlation of thermal sputtering and the number of atoms in the volume effected by the ion, because a comparable amount of energy is distributed amongst fewer atoms in thinner nanowires than in thicker ones.

That thermal effects certainly play a significant role is illustrated by the plastic deformation that $Si$-nanowires show when irradiated with medium weight $Ar^+$ ions at energies of $100\,keV$ and room-temperature and not at elevated temperatures. It could be shown that this deformation is not mediated by point defects and is not directed along the ion beam. Where the lower ion energy threshold for the deformation is and whether there is an upper threshold ion energy above which the deformation ceases is not clear. The deformation is observed in amorphous $Si$-nanowires, but not in crystalline $Si$, even if both are irradiated at elevated temperatures. In the crystalline case, the efficient recrystallization of the ion damaged nanowire volume recreates the long range order of the crystal lattice after every ion impact. The amorphous material remains incrementally deformed, even if it is irradiated at elevated temperatures. This has also been observed in the plastic deformation or viscous flow of bulk samples by swift-heavy ion irradiation, which can be explained very well by ion track induced plastic deformation as proposed by Trinkaus et al. in reference \cite{trinkaus_viscoelastic_1995}. It is clear, however, that the observed deformation in the experiments reported in this thesis is not another example of the mechanism reported by Trinkaus, because the energy loss of the ion is too low to form any sort of ion track. Also, it is not certain how the locally increased pressure required by an adaptation of this model can be confined to the limited volume of a nanowire. Therefore, a surface tension driven model, which relies on a locally reduced viscosity, was presented in this dissertation. This model is further supported by the pronounced smoothing of the deformed nanowires' surface and edges, which is especially visible at the rounded top facet, which the crystalline, undeformed nanowires irradiated at elevated temperatures do not show.

Regardless of the underlying mechanism, the plastic deformation of $Si$ has great potential for nanostructuring applications, because it is highly localized at the point of the ions impact. In combination with the relatively low ion energy required it is predestined to be used in FIB assisted fabrication of nanostructures. It may be relevant to the formation of nanopores \cite{george_nanopore_2010} and certainly to the bending and manipulation of nanowires \cite{cui_ion-beam-induced_2013} and freestanding films \cite{kim_focused_2006}. With suitable templates that may have to be thinned at the bending points, it could be possible to go as far as building `$Si$-nano-origami' similar to the metal `origami' fabricated in reference \cite{chalapat_self-organized_2013}. Such folding of $Si$-nanoshapes in particular may be a versatile fabrication tool in the growing field of $Si$ MEMS devices. Furthermore, plastic deformation may have to be considered in the formulation of the mechanism for the formation of ripples on ion irradiated $Si$ surfaces. The dated model by Bradley and Harper \cite{bradley_theory_1988} considers only curvature and angle dependent sputtering as a roughening mechanism and is contested by models including ion induced strain and mass-transport \cite{norris_stress-induced_2012,kramczynski_wavelength-dependent_2014}. The latter shows some similarity with the results presented here, indicating that an atomistic investigation may be necessary to resolve this issue.

All three chapters of this thesis have compared MC BCA simulations performed with \emph{iradina} to experimental results on nanowires irradiated with high ion fluences. One limit to the applicability of the BCA is found were thermal effects have to be considered. This is not quite the case for the presented diameter-dependent sputter yields, where the simulation overlaps qualitatively with the experimental results. However, the plastic deformation found in amorphous $Si$-nanowires can not be explained with this simulation technique. Furthermore, the obtainable accuracy of \emph{iradina} simulations of the prediction of the doping concentration in nanostructures determined with is satisfactory, but limited to low ion fluences. When the desired doping concentration is high and high ion fluences have to be implanted, dynamic simulations become neccessary. A rule of thumb lower limit to what constitutes a `high' ion fluence is given by the fluence at which $20\%$ of the volume effected by the ion beam is sputtered. These limitations do not formulate a criticism of the program \emph{iradina} or the MC BCA simulation technique in general, but only serve to warn against their naive use in situations in which they are not applicable.
 


